---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MultiMapR

<!-- badges: start -->
<!-- badges: end -->

**MultiMapR** is an R package for mapping and visualizing discrete characters on phylogenies. It projects one or more morphological, ecological, or molecular characters directly onto a tree — either at the terminals or through ancestral state reconstruction — with support for three topologies (phylogram, cladogram, fan) and export to publication-ready PNG and PDF files.

## Installation

You can install the development version of MultiMapR from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("your-username/MultiMapR")
```

MultiMapR requires the `ape` package (>= 5.0), which is installed automatically as a dependency.

## Quick start

```{r example, eval = FALSE}
library(MultiMapR)

# Option A — pass file paths directly (recommended)
execute_phylogeny("my_tree.tre", "my_characters.csv")

# Option B — load objects first, then map
data <- load_data("my_tree.tre", "my_characters.csv")
execute_phylogeny(data$tree, data$characters)
```

Calling `execute_phylogeny()` opens an **interactive console menu** that guides you through every option: mapping type, colours per state, algorithm, tree topology, and export settings. Type `exit` at any prompt to cancel.

## Loading data

`load_data()` handles tree and character matrix reading in a single call:

```{r load-data, eval = FALSE}
# Basic usage — auto-detects format, header, and BOM
d <- load_data("tree.tre", "characters.csv")

# Semicolon-separated CSV, species column identified by name
d <- load_data("tree.nex", "data.csv",
               sep = ";", species_col = "Taxon")

# Normalise spaces to underscores when tree uses "_" but CSV uses " "
d <- load_data("tree.tre", "data.csv", normalize_spaces = TRUE)

# Raise an error (instead of a warning) on tree/CSV mismatches
d <- load_data("tree.tre", "data.csv", strict = TRUE)
```

`load_data()` returns a list with two elements:

- `$tree` — a `phylo` object ready for ape and MultiMapR.
- `$characters` — a data frame with a `Species` column and one column per character.

### Supported input formats

**Phylogenetic tree**

| Extension | Format | Parser |
|---|---|---|
| `.tre`, `.tree`, `.nwk` | Newick | `ape::read.tree()` |
| `.nex`, `.nexus` | NEXUS | `ape::read.nexus()` |

Format is detected automatically from the file extension. Mesquite-style numeric comments (`[0]`, `[1]`…) and Windows line endings (`\r\n`) are cleaned before parsing.

**Character matrix (CSV)**

- First column: species names (header auto-detected).
- Remaining columns: discrete characters (text or numbers).
- Configurable separator (default `,`).
- Excel UTF-8 BOM removed automatically.
- Inapplicable (`-`) and unknown (`?`) states preserved and rendered in grey.
- Polymorphisms supported: `"0/1"`, `"urban,forest"`, `"01"` (numeric without separator).

Example CSV **with** header:

```
Species,diet,locomotion,habitat
Homo_sapiens,omnivore,bipedalism,terrestrial
Pan_troglodytes,omnivore,quadrupedalism,arboreal
Gorilla_gorilla,herbivore,quadrupedalism,terrestrial
```

Example CSV **without** header:

```
Homo_sapiens,0,1,2
Pan_troglodytes,0,0,1
Gorilla_gorilla,1,0,0
```

## Mapping modes

### Simple mapping

Draws the tree in neutral grey and places **coloured symbols** (circle, square, triangle, or diamond) at each terminal — one column per character. Ideal for comparing state distributions without ancestral inference.

- Supports **one or more characters simultaneously**.
- **Fan** topology: concentric rings of symbols around the tree.
- **Phylogram / cladogram**: symbol table to the right of tip labels.

### Ancestral reconstruction — single character

Colours the **branches** of the tree according to the state inferred at each internal node, using the algorithm chosen in the menu.

### Ancestral reconstruction — multi-character (up to 3)

Superimposes the reconstruction of several characters on the same tree using **isotropic offsets** (equal in X and Y, computed at render time from the actual canvas dimensions) so that each evolutionary history remains readable without overlap.

## Ancestral reconstruction algorithms

### Default algorithm (depth-weighted majority)

Traverses internal nodes in post-order and assigns each edge the colour that predominates among its descendants, prioritising the closest internal nodes in the hierarchy. Fast, requires no extra parameters, and works well for trees with many states.

### Fitch algorithm

Full implementation of the Fitch parsimony algorithm (Swofford & Maddison, 1987) with three ambiguity-resolution modes:

| Mode | Behaviour |
|---|---|
| **ACCTRAN** | Uses descending-pass sets for ambiguous nodes. Accelerates changes towards the tips (places them as early as possible). |
| **DELTRAN** | Uses MPR sets from the ascending pass. Delays changes towards the root (places them as late as possible). |
| **Unambiguous** | A node receives a state only when ACCTRAN and DELTRAN agree exactly. All others remain grey. |

The algorithm handles polytomies, polymorphic states, and missing data out of the box.

## Interactive menu (console walkthrough)

```
=== Welcome to MultiMapR ===

Mapping type:
  1: Simple mapping — coloured symbols at terminals
  2: Ancestral reconstruction — branch colouring
Select (1/2, or 'exit' to quit): _

[ancestral reconstruction]

Visualisation mode:
  1: Branch superimposition (one or more characters)
  2: Terminal symbols + coloured tree

Reconstruction algorithm:
  1: Default (depth-weighted majority)
  2: Fitch (ACCTRAN / DELTRAN / Unambiguous)

Fitch optimisation mode:
  1: ACCTRAN     (accelerated changes — towards tips)
  2: DELTRAN     (delayed changes — towards root)
  3: Unambiguous (only unambiguous nodes coloured)

Branch width (positive number; Enter = 2):
  Reference: thin ≈ 1, normal ≈ 2, thick ≈ 4, very thick ≈ 8

Tree type: phylogram / cladogram / fan

Export the figure?
  1: Yes — save as file
  2: No — display in R window

Output format:
  1: PNG (recommended for screen / presentations)
  2: PDF (vector, ideal for publications)
```

## Export

Exported files are saved to an `Exports/` folder in the current working directory, created automatically if it does not exist.

Default dimensions scale with tip count:

```
height  = n_tips × 0.25 + 2  (inches)
width   = 12                  (inches)
resolution = 300 dpi          (PNG only)
```

Fan trees use equal width and height. When custom dimensions are specified, the text size (`cex`) is rescaled proportionally so labels never appear too small or overlap.

## Package architecture

```
MultiMapR/
│
├── R/
│   ├── main_MultiMapR.R         ← Orchestrator: execute_phylogeny()
│   ├── load_data.R              ← Data reading and validation
│   ├── cli_menu.R               ← Interactive console menus
│   ├── data_utils.R             ← Pure utilities (colours, alignment, cex)
│   ├── core_render.R            ← Rendering engines and reconstruction
│   ├── export_multimapr_tree.R  ← Universal export (PNG / PDF)
│   ├── default_alg.R            ← Default reconstruction algorithm
│   └── fitch.R                  ← Fitch algorithm (ACCTRAN/DELTRAN/Unambiguous),
│                                   exported as external_algorithm()
│
└── DESCRIPTION                  ← Imports: ape (>= 5.0), tools
```

Each module has a single responsibility. `core_render.R` never touches disk; `load_data.R` never draws; `cli_menu.R` only handles user input. Private helpers carry a `.` prefix and are not exported to the namespace.

### Internal dispatch

`core_render.R` dispatches ancestral reconstruction via `apply_ancestral_algorithm()`:

| `config$algoritmo` | Function called | Config key used |
|---|---|---|
| `1` | `default_algorithm(tree, tip_colors, edge_colors)` | — |
| `2` | `external_algorithm(tree, tip_colors, edge_colors, config)` | `config$fitch_mode` |

`config$fitch_mode` is a string (`"acctran"`, `"deltran"`, or `"unambiguous"`) set by `cli_menu.R`.

## Function reference

| Function | Description |
|---|---|
| `execute_phylogeny(tree, data, ...)` | Main orchestrator. Accepts file paths or in-memory objects. |
| `load_data(tree_path, csv_path, ...)` | Loads and validates tree + character matrix. |
| `read_tree(path, format = "auto")` | Reads a tree file independently. |
| `read_csv_characters(path, ...)` | Reads a character CSV independently. |
| `validate_compatibility(tree, data)` | Checks tip-name/species-name agreement. |

## FAQ

**What R version is required?**
R ≥ 4.1.0 and `ape` ≥ 5.0.

**My species names have spaces in the CSV but underscores in the tree.**
Use `load_data(..., normalize_spaces = TRUE)` to convert spaces to `_` automatically before matching.

**How many characters can I map at once?**
Simple mapping has no limit. Ancestral reconstruction supports up to **3 characters** to keep the superimposition readable.

**The figure looks fine on screen but tip labels are tiny in the PNG.**
Increase the height in inches via the custom dimensions menu. MultiMapR rescales `cex` automatically to match your chosen canvas size.

**Fitch Unambiguous mode shows many grey branches.**
That is the intended behaviour: only nodes where ACCTRAN and DELTRAN agree receive a colour. Use ACCTRAN or DELTRAN if you need full branch coverage.

**Where are exported files saved?**
In an `Exports/` folder inside your current working directory (`getwd()`).

## Reference

Swofford, D. L. & Maddison, W. P. (1987). Reconstructing ancestral character states under Wagner parsimony. *Systematic Zoology*, 36(4), 367–379.
